---
title: 'Assessment Number 1: Portfolio 2 of 4'
author: "C.U.V.A."
student ID: Chiamaka Chijioke (2423904), 
            Udochukwu Okike (2423893), 
            Victor Ene (2318217), 
            Amarachi Eche (2423917)
date: "2025-03-20"
---

# 1.0 Library and Package Installation

```{r Install and Load Packages}
# Define a set of required packages for data analysis, web scraping, and visualization
packages <- c("tidyverse", "rvest", "xml2", "httr", "skimr", "reshape2")

# Install any packages that are not already installed
install.packages(setdiff(packages, installed.packages()[, "Package"]), dependencies = TRUE)

# Load the necessary libraries
library(tidyverse)    # For data manipulation and visualizationIncludes dplyr, ggplot2, readr, stringr, purrr, etc.
library(rvest)        # For web scraping (HTML parsing)
library(xml2)         # Backend for rvest, parses XML/HTML documents
library(httr)         # For HTTP requests
library(skimr)        # For generating summary statistics
library(reshape2)     # For correlation heatmap
```

# 2.0 Site 1: CAR SITE

## 2.1 Webscraping

```{r CarSite Scraping}
# Function to generate the URL for a specific page number on CarSite
get_carsite_url <- function(page_num) {
  paste0("https://www.carsite.co.uk/used-car/page/", page_num)  # Concatenate base URL with page number
}

# Safely extract text from an HTML node, returning NA if the selector fails
safe_extract <- function(node, selector) {
  result <- node %>% html_element(selector) %>% html_text(trim = TRUE)  # Extract text and trim whitespace
  if (length(result) == 0) return(NA) else return(result)  # Return NA if no result, otherwise the text
}

# Function to scrape car data from a single CarSite page
scrape_carsite_page <- function(page_num) {
  url <- get_carsite_url(page_num)  # Generate URL for the page
  cat("Scraping Page:", page_num, "\n")  # Print progress message
  response <- GET(url, user_agent("CUVA for Educational Use")) # For client identification 
  page <- tryCatch(read_html(url), error = function(e) NULL)  # Attempt to read the page, return NULL on error
  if (is.null(page)) return(NULL)  # Exit if page loading failed
  
  cars <- page %>% html_elements(".paContainer")  # Select all car listing containers
  
  # Map over each car listing to extract data into a data frame
  car_data <- map_df(cars, function(car) {
    data.frame(
      Car_Name = safe_extract(car, "h3.avlink a"),              # Car name from the link title
      Price = safe_extract(car, "h4.advertprice"),             # Price from the advert
      Mileage = safe_extract(car, "div.section28.fl:nth-child(2)"),  # Mileage from second div
      Transmission = safe_extract(car, "div.section28.fl.mr1p:nth-child(1)"),  # Transmission type
      Fuel_Type = safe_extract(car, "div.section28.fl:nth-child(4)"),  # Fuel type
      Engine_Size = safe_extract(car, "div.section28.fl.mr1p:nth-child(5)"),  # Engine size
      Doors = safe_extract(car, "div.section28.fl:nth-child(6)"),  # Number of doors
      Color = safe_extract(car, "div.section28.fl.mr1p:nth-child(3)"),  # Car color
      Year_Model = safe_extract(car, ".section.f-12.pos-rel span.tal")  # Year and model info
    )
  })
  Sys.sleep(runif(1, 2, 5))  # Random delay between 2-5 seconds to avoid overloading the server
  return(car_data)  # Return the scraped data
}

# Scrape data from pages 1 to 300 and combine into a single data frame
carsite_raw <- map_df(1:300, scrape_carsite_page)
View(carsite_raw)  # Display the raw scraped data
```

## 2.2 Data Preprocessing

```{r}
# Tidy and preprocess the raw CarSite data
carsite_data <- carsite_raw %>%
  mutate(Price = parse_number(Price),             # Convert Price to numeric, removing non-numeric characters
         Mileage = parse_number(Mileage),         # Convert Mileage to numeric
         Year = as.numeric(str_extract(Year_Model, "^\\s*(\\d{4})")),           # Ensure Year is numeric
         Fuel_Type = if_else(Fuel_Type == "-", "Hybrid", Fuel_Type), # Replace "-" in Fuel_Type with "Hybrid"
         Engine_Size = if_else(Engine_Size == "-", NA_character_, Engine_Size), # Replace "-" with NA
         Engine_Size = parse_number(Engine_Size)) %>%  # Convert Engine_Size to numeric
  extract(Year_Model, into = c("Year", "Make", "Model", "Car_Type"),  # Split Year_Model into components
          regex = "^\\s*(\\d{4})\\s+((?:Land\\s+Rover)|\\S+)\\s+(.+?)\\s+(\\S+)\\s*$",
          remove = FALSE) %>%  # Use regex to extract Year, Make, Model, and Car_Type
 
# Select and rename columns for consistency and clarity
  select(Car_Brand = Make,                # Rename Make to Car_Brand
         Car_Model = Model,              # Rename Model to Car_Model
         `Price (£)` = Price,            # Rename Price with units
         `Mileage (miles)` = Mileage,    # Rename Mileage with units
         `Engine_Size (cc)` = Engine_Size,  # Rename Engine_Size with units
         Fuel = Fuel_Type)               # Rename Fuel_Type to Fuel

print(tail(carsite_data, 10))  # Show sample
```

# 3.0 Site 2: Autouncle

## 3.1 Webscrapping

```{r}
# Define base URL and parameters for scraping Autouncle
base_url <- "https://www.autouncle.co.uk/en-gb/used-cars?page="
start_page <- 1
num_pages <- 100

# Function to extract text from an HTML node, returning NA if empty
extract_text <- function(node, selector) {
  node %>% html_element(selector) %>% html_text(trim = TRUE) %>% ifelse(. == "", NA, .)  # Extract and handle empty results
}

# Function to scrape car data from a single Autouncle page
scrape_page <- function(url) {
  page <- tryCatch(read_html(url), error = function(e) NULL)  # Attempt to read the page, return NULL on error
  if (is.null(page)) return(NULL)  # Exit if page loading failed
  
  # Extract car listings and map to a data frame
  page %>% html_elements(".styles_listing-item__KWJkL") %>%
    map_df(~ data.frame(
      Make_Model = extract_text(.x, ".styles_headline__TnlnI"),  # Combined make and model
      Price = extract_text(.x, ".styles_price__X069_"),         # Price of the car
      Mileage = extract_text(.x, ".styles_label__x4zE1:nth-child(2) .styles_text__eMFHA"),  # Mileage
      Year = extract_text(.x, ".styles_highlighted-attributes-container__EH8XY .styles_label__x4zE1:nth-child(1) .styles_text__eMFHA"),  # Year
      Transmission = extract_text(.x, ".styles_label__x4zE1:nth-child(3) .styles_text__eMFHA"),  # Transmission
      Fuel_Type = extract_text(.x, ".styles_highlighted-attributes-container__EH8XY .styles_label__x4zE1:nth-child(4) .styles_text__eMFHA"),  # Fuel type
      Power = extract_text(.x, ".styles_label__x4zE1:nth-child(6) .styles_text__eMFHA"),  # Engine power
      stringsAsFactors = FALSE  # Prevent factor conversion
    ))
}

# Scrape data from 100 pages and combine into a single data frame
all_cars <- map_df(start_page:(start_page + num_pages - 1), function(page_num) {
  url <- paste0(base_url, page_num)  # Construct URL for the page
  cat("Scraping page:", page_num, "\n")  # Print progress message
  sleep_time <- runif(1, 2, 4)  # Random delay between 2-4 seconds
  if (runif(1) < 0.1) { Sys.sleep(runif(1, 5, 10)) }  # Occasionally add a longer delay (5-10 seconds)
  Sys.sleep(sleep_time)  # Apply the delay
  scrape_page(url)  # Scrape the page
})
View(all_cars)
```

## 3.2 Data Preprocessing

```{r}
# Function to clean Mileage by removing non-numeric characters
clean_mileage <- function(data) {
  data %>%
    mutate(Mileage = parse_number(Mileage))
}

# Function to split Fuel_Type and extract Engine_Size
split_fuel_engine <- function(data) {
  data %>%
    mutate(
      Engine_Size = ifelse(str_detect(Fuel_Type, "\\d\\.\\dL"),  # Check if Fuel_Type contains engine size (e.g., 2.0L)
                           str_extract(Fuel_Type, "\\d\\.\\dL"),  # Extract engine size
                           "0.0"),  # Default to 0.0 if not found(Electric cars don't have Engines)
      Fuel = str_remove(Fuel_Type, "\\d\\.\\dL\\s*"),  # Remove engine size from Fuel_Type
      Fuel = str_trim(Fuel)  # Trim whitespace
    )
}

# Function to convert Engine_Size from liters to cubic centimeters (cc)
convert_engine_to_cc <- function(data) {
  data %>%
    mutate(
      Engine_Size = as.numeric(str_remove(Engine_Size, "L")),  # Remove "L" and convert to numeric
      `Engine_Size (cc)` = Engine_Size * 1000  # Convert liters to cc
    ) %>%
    select(-Engine_Size)  # Drop the original Engine_Size column
}

# Function to clean Price by removing currency symbols and commas
clean_price <- function(data) {
  data %>%
    mutate(
      `Price (£)` = as.numeric(str_remove_all(Price, "[£,]"))  # Remove £ and commas, convert to numeric
    ) %>%
    select(-Price)  # Drop the original Price column
}

# Function to extract Car Type from Make_Model using a predefined list
extract_car_type <- function(data) {
  car_types <- c("Saloon", "Hatchback", "SUV", "Sports Utility Vehicle",
                 "Crossover", "Coupe", "Coupé", "Convertible", "Cabriolet",
                 "Estate", "Pick[- ]?up(?: Truck)?", "Minivan", "MPV",
                 "Roadster", "Van", "Cargo Van", "City Car", "Microcar",
                 "Liftback", "Fastback", "Targa(?: Top)?", "Limousine",
                 "Supercar", "Hypercar", "4x4", "Off[- ]?Roader", "Muscle Car",
                 "Coupe SUV", "Campervan", "Recreational Vehicle", "RV",
                 "Sportback", "Panel Van", "Combi Van", "Touring",
                 "Shooting Brake", "Double Cab", "Quad Cab",
                 "Extended Cab", "Supercrew", "Hardtop", "Softtop")
  
  car_type_pattern <- paste0("(?i)(", paste(car_types, collapse = "|"), ")")  # Create regex pattern for car types
  
  data %>%
    mutate(`Car Type` = str_extract(Make_Model, car_type_pattern),  # Extract car type from Make_Model
           `Car Type` = str_to_title(str_trim(`Car Type`)))  # Capitalize and trim
}

# Function to extract Car Brand from Make_Model using a predefined list
extract_car_brand <- function(data) {
  brands <- c("Abarth", "AC", "Acura", "Aixam", "Alfa Romeo", "Alpine", 
              "Aston Martin", "Audi", "Austin", "Bentley", "BMW", "Bugatti", 
              "Buick", "Cadillac", "Caterham", "Chevrolet", "Chrysler", 
              "Citroën", "Cupra", "Dacia", "Daewoo", "Daihatsu", "Dodge", 
              "DS Automobiles", "Eagle", "Ferrari", "Fiat", "Fisker", 
              "Ford", "Genesis", "GMC", "Great Wall", "Haval", "Holden", 
              "Honda", "Hummer", "Hyundai", "Infiniti", "Isuzu", "Jaguar", 
              "Jeep", "Kia", "Lada", "Lamborghini", "Lancia", "Land Rover", 
              "Lexus", "Lincoln", "Lotus", "Mahindra", "Maserati", "Maybach", 
              "Mazda", "McLaren", "Mercedes", "MG", "Mini", "Mitsubishi", 
              "Nissan", "Oldsmobile", "Opel", "Pagani", "Peugeot", "Plymouth", 
              "Polestar", "Pontiac", "Porsche", "Proton", "Ram", "Range Rover", 
              "Renault", "Rolls-Royce", "Rover", "Saab", "Seat", "Skoda", 
              "Smart", "SsangYong", "Subaru", "Suzuki", "Tata", "Tesla", 
              "Toyota", "Vauxhall", "Volkswagen", "VW", "Volvo", "Wiesmann", "Zotye")
  brands <- brands[order(nchar(brands), decreasing = TRUE)]  # Sort brands by length (longest first)
  
  data %>%
    mutate(
      Car_brand = sapply(Make_Model, function(make_model) {  # Apply function to each Make_Model
        brand_found <- NA
        for (brand in brands) {  # Loop through brands
          if (str_starts(make_model, brand)) {  # Check if Make_Model starts with the brand
            brand_found <- brand
            break
          }
        }
        return(brand_found)  # Return the matched brand or NA
      }),
      Make_Model = str_trim(ifelse(is.na(Car_brand), Make_Model,  # Remove brand from Make_Model if found
                                   str_remove(Make_Model, paste0("^", Car_brand))))
    )
}

# Function to extract Car Model from remaining Make_Model text
extract_car_model <- function(data) {
  data %>%
    mutate(
      Car_model = if_else(
        str_detect(Make_Model, "\\s\\d\\.\\d"),  # Check for engine size pattern (e.g., "2.0")
        str_trim(str_extract(Make_Model, ".*?(?=\\s\\d\\.\\d)")),  # Extract text before engine size
        str_trim(Make_Model)  # Otherwise use the full remaining text
      )
    )
}

# Function to clean Year by extracting a 4-digit year
clean_year <- function(data) {
  data %>%
    mutate(
      Year = case_when(
        str_detect(Year, "\\b\\d{4}\\b") ~ str_extract(Year, "\\b\\d{4}\\b"),  # Extract 4-digit year
        TRUE ~ NA_character_  # Return NA if no match
      ),
         Year = as.numeric(Year)  # Convert to numeric
    )
}

# Function to further clean Car_model by removing unwanted terms
final_clean_car_model <- function(data) {
  data %>%
    mutate(
      Car_model = str_split(Car_model, "\\s+") %>%  # Split into words
        lapply(function(words) {
          words_clean <- words[!str_detect(tolower(words), "dr|kwh|kw|auto")]  # Remove specific terms
          paste(words_clean, collapse = " ")  # Rejoin words
        }) %>%
        unlist() %>%  # Convert back to vector
        str_squish()  # Remove extra whitespace
    )
}

# Execute the cleaning pipeline for Autouncle data
auto_cars <- all_cars %>%
  clean_mileage() %>%              # Clean mileage values
  split_fuel_engine() %>%          # Split fuel and engine size
  convert_engine_to_cc() %>%       # Convert engine size to cc
  clean_price() %>%                # Clean price values
  extract_car_type() %>%           # Extract car type
  extract_car_brand() %>%          # Extract car brand
  extract_car_model() %>%          # Extract car model
  clean_year() %>%                 # Clean year values
  final_clean_car_model() %>%      # Final cleanup of car model
  select(-Make_Model, -Fuel_Type, -Power)  # Drop unnecessary columns

head(auto_cars, 10)  # Display the cleaned Autouncle data
```

# 4.0 Merging Datasets Handling Missing Values and Detecting Outliers

```{r}

# Align column names of Autouncle data with CarSite data
auto_cars_aligned <- auto_cars %>%
  rename(
    Car_Brand = Car_brand,              # Rename to match CarSite
    Car_Model = Car_model,              # Rename to match CarSite
    Car_Type = `Car Type`,              # Rename to match CarSite
    `Mileage (miles)` = Mileage         # Rename to match CarSite
  ) %>%
  select(
    Car_Brand, Car_Model, Year, Car_Type, `Price (£)`, `Mileage (miles)`, Transmission, Fuel, `Engine_Size (cc)`
  )  # Select and order columns to match CarSite

# Combine the two datasets and standardize column names
Used_cars <- bind_rows(carsite_data, auto_cars_aligned) %>%
  distinct() %>%  # Remove duplicates
  rename(
    Price_in_GBP = `Price (£)`,           # Rename with more descriptive name
    Mileage_in_miles = `Mileage (miles)`, # Rename with more descriptive name
    Engine_Size_in_cc = `Engine_Size (cc)` # Rename with more descriptive name
  ) %>%
  select(Car_Brand, Car_Model, Year, Car_Type, Price_in_GBP, Mileage_in_miles, Transmission, Fuel, Engine_Size_in_cc) %>% # Reorder columns
  drop_na()  # Remove rows with any missing values

View(Used_cars)  # Display the merged dataset
skim(Used_cars)  # Generate summary statistics
write.csv(Used_cars, "Used_cars.csv", row.names = FALSE)  # Export the merged dataset to a CSV file
```

## 4.1 Outlier Analysis

```{r}
# Define outlier detection function
detect_outliers_iqr <- function(x) {
  q <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
  iqr <- diff(q)
  x < q[1] - 1.5 * iqr | x > q[2] + 1.5 * iqr
}

#Quick outlier check using IQR (Detection only)
outlier_summary <- Used_cars %>%
  summarise(across(c(Price_in_GBP, Mileage_in_miles, Engine_Size_in_cc),
                   ~sum(detect_outliers_iqr(.)), .names = "{col}_outliers")) %>%
  print()

# Example high-price outliers
high_price_outliers <- Used_cars %>%
  filter(detect_outliers_iqr(Price_in_GBP)) %>%
  arrange(desc(Price_in_GBP)) %>%
  select(Car_Brand, Car_Model, Year, Price_in_GBP, Mileage_in_miles, Engine_Size_in_cc) %>%
  head() %>%
  print()

# Decision
cat("Outliers retained for EDA to reflect market diversity (e.g., luxury cars, high-mileage vehicles).\n")
```

# 5.0 EDA

##### Grouped Data Summary Group by Car_Brand

```{r}
# Summarize data by Car_Brand with price statistics
grouped_Car_Brand <- Used_cars %>%
  group_by(Car_Brand) %>%  # Group by car brand
  summarize(Average_Price = round(mean(Price_in_GBP)),  # Calculate mean price
            Median_Price = round(median(Price_in_GBP)),  # Calculate median price
            Min_Price = min(Price_in_GBP),        # Calculate minimum price
            Max_Price = max(Price_in_GBP),        # Calculate maximum price
            Count = n()) %>%                      # Count number of cars per brand
  arrange(desc(Count))  # Sort by count in descending order
print(grouped_Car_Brand)  # Display the summary
```

##### Group by Fuel

```{r}
# Summarize data by Fuel type with price statistics
grouped_Fuel <- Used_cars %>%
  group_by(Fuel) %>%  # Group by fuel type
  summarize(Average_Price = round(mean(Price_in_GBP)),  # Calculate mean price
            Median_Price = round(median(Price_in_GBP)),  # Calculate median price
            Min_Price = min(Price_in_GBP),        # Calculate minimum price
            Max_Price = max(Price_in_GBP),        # Calculate maximum price
            Count = n()) %>%                      # Count number of cars per fuel type
  arrange(desc(Count))  # Sort by count in descending order
print(grouped_Fuel)  # Display the summary
```

##### Group by Transmission

```{r}
# Summarize data by Transmission type with price statistics
grouped_Transmission <- Used_cars %>%
  group_by(Transmission) %>%  # Group by transmission type
  summarize(Average_Price = round(mean(Price_in_GBP)),  # Calculate mean price
            Median_Price = round(median(Price_in_GBP)),  # Calculate median price
            Min_Price = min(Price_in_GBP),        # Calculate minimum price
            Max_Price = max(Price_in_GBP),        # Calculate maximum price
            Count = n()) %>%                      # Count number of cars per transmission
  arrange(desc(Count))  # Sort by count in descending order
print(grouped_Transmission)  # Display the summary
```

##### Group by Car_Type

```{r}
# Summarize data by Car_Type with price statistics
grouped_Car_Type <- Used_cars %>%
  group_by(Car_Type) %>%  # Group by car type
  summarize(Average_Price = round(mean(Price_in_GBP)),  # Calculate mean price
            Median_Price = round(median(Price_in_GBP)),  # Calculate median price
            Min_Price = min(Price_in_GBP),        # Calculate minimum price
            Max_Price = max(Price_in_GBP),        # Calculate maximum price
            Count = n()) %>%                      # Count number of cars per car type
  arrange(desc(Count))  # Sort by count in descending order
print(grouped_Car_Type)  # Display the summary
```

## 5.1 Function to Calculate summary statistics

```{r}
# Calculate summary statistics for Price_in_GBP
price_summary <- Used_cars %>%
  summarise(
    Mean = mean(`Price_in_GBP`, na.rm = TRUE),      # Calculate mean price
    Median = median(`Price_in_GBP`, na.rm = TRUE),  # Calculate median price
    SD = sd(`Price_in_GBP`, na.rm = TRUE),          # Calculate standard deviation
    Min = min(`Price_in_GBP`, na.rm = TRUE),        # Calculate minimum price
    Q1 = quantile(`Price_in_GBP`, 0.25, na.rm = TRUE),  # Calculate first quartile
    Q3 = quantile(`Price_in_GBP`, 0.75, na.rm = TRUE),  # Calculate third quartile
    Max = max(`Price_in_GBP`, na.rm = TRUE)         # Calculate maximum price
  )
print(price_summary)  # Display the summary statistics
```

## 5.2 Creating visualizations using Histogram for Price

```{r}
# Create a histogram to visualize the distribution of car prices
ggplot(Used_cars, aes(x = `Price_in_GBP`)) +
  geom_histogram(binwidth = 5000, fill = "blue", color = "black") +  # Plot histogram with 5000 GBP bins
  labs(title = "Price Distribution", x = "Price (£)", y = "Frequency")  # Add titles and labels
```

## 5.3 Box Plot for Price

```{r}

# Create box plots to visualize the spread and outliers of car prices, mileages and engine size
Used_cars %>%
  pivot_longer(cols = c(Price_in_GBP, Mileage_in_miles, Engine_Size_in_cc),
               names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(y = Value)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +  # Plot box plot with custom colors
  facet_wrap(~Variable, scales = "free_y") +
  labs(title = "Box Plots for Numeric Variables", y = "Value") # Add title and y-axis label
```

## 5.4 Function to generate a scatterplot with trend line

```{r}
# Function to create a scatter plot with a trend line for two variables
scatter_plot <- function(Used_cars, x_var, y_var) {
  Used_cars %>%
    ggplot(aes_string(x = x_var, y = y_var)) +  # Use aes_string for dynamic column names
    geom_point(color = "blue", alpha = 0.05) +   # Plot points with transparency
    geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add linear trend line without confidence interval
    labs(title = paste("Scatterplot of", y_var, "vs", x_var),  # Dynamic title
         x = x_var, y = y_var) +                       # Axis labels
    theme_minimal()  # Use minimal theme
}

# Generate scatter plots for key relationships
Used_cars %>% scatter_plot("Mileage_in_miles", "Price_in_GBP")  # Mileage vs Price
Used_cars %>% scatter_plot("Year", "Price_in_GBP")             # Year vs Price
Used_cars %>% scatter_plot("Year", "Mileage_in_miles")         # Year vs Mileage
Used_cars %>% scatter_plot("Engine_Size_in_cc", "Price_in_GBP")  # Engine Size vs Price


```

### 5.5 Function to generate a correlation heatmap for numeric variables in a data frame

```{r}

# Function to create a correlation heatmap for numeric variables
correlation_heatmap <- function(data) {
    # Calculate the correlation matrix (using complete observations) and Select only numeric columns
  corr_matrix <- cor(select_if(data, is.numeric), use = "complete.obs")
    # Convert the correlation matrix to long format for ggplot2 and create the heatmap using ggplot2
  ggplot(reshape2::melt(corr_matrix), aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() + # Plot tiles for each correlation value
    scale_fill_gradient2(low = "blue", high = "red", mid = "white",  # Color gradient from -1 to 1
                         midpoint = 0, limit = c(-1, 1)) +
    labs(title = "Correlation Heatmap", x = "", y = "") +  # Add title, remove axis labels
    theme_minimal() + # Use minimal theme
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) # Rotate x-axis labels
}

# Generate the correlation heatmap for the main data frame
correlation_heatmap(Used_cars)

```