---
title: "DAT301_As2"
author: "Udochukwu Okike 2423983"
date: "2025-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Packages

```{r packages}
# Install packages if not already installed
required_packages <- c(
  "tidyverse", "readxl", "lubridate", "janitor", "ggplot2", "ggpubr",
  "scales", "reshape2", "forecast", "tseries", "randomForest", "multcomp",
  "caret", "cluster", "factoextra", "patchwork", "corrplot", "tibble"
)

installed_packages <- rownames(installed.packages())
for (pkg in required_packages) {
  if (!(pkg %in% installed_packages)) {
    install.packages(pkg)
  }
}

# Load packages
lapply(required_packages, library, character.only = TRUE)
```

# Loading, Reading and Tidying The Dataset

## Load and Inspect Data by Year

```{r load-data}
# Read data directly from Excel sheets with cleaned column names using janitor::clean_names()
df_2019_raw <- read_excel("IGR_DATA_2019_2023.xlsx", sheet = "2019-2021", range = "B2:E39") %>% clean_names()
df_2020_raw <- read_excel("IGR_DATA_2019_2023.xlsx", sheet = "2019-2021", range = "F2:H39") %>% clean_names()
df_2021_raw <- read_excel("IGR_DATA_2019_2023.xlsx", sheet = "2019-2021", range = "I2:K39") %>% clean_names()
df_2022_raw <- read_excel("IGR_DATA_2019_2023.xlsx", sheet = "2022", range = "B2:M39") %>% clean_names()
df_2023_raw <- read_excel("IGR_DATA_2019_2023.xlsx", sheet = "2023", range = "B5:M42") %>% clean_names()
```

## Extract and Combine State, Year and Total

```{r tidy-combined-minimal}
states <- df_2019_raw$x1 %>% tolower()

# Enhanced cleaning function for known inconsistencies
clean_state <- function(x) {
  x <- tolower(x)
  x <- gsub("federal capital territory", "fct", x)
  x <- gsub("kaduna\\*", "kaduna", x)
  x <- gsub("nassarawa", "nasarawa", x)
  return(x)
}

df_2019 <- df_2019_raw %>% mutate(state = clean_state(x1), year = 2019, total = as.numeric(total)) %>% dplyr::select(state, year, total)
df_2020 <- df_2020_raw %>% mutate(state = clean_state(states), year = 2020, total = as.numeric(total)) %>% dplyr::select(state, year, total)
df_2021 <- df_2021_raw %>% mutate(state = clean_state(states), year = 2021, total = as.numeric(total)) %>% dplyr::select(state, year, total)
df_2022 <- df_2022_raw %>% mutate(state = clean_state(state), year = 2022, total = as.numeric(total)) %>% dplyr::select(state, year, total)
df_2023 <- df_2023_raw %>% mutate(state = clean_state(state), year = 2023, total = as.numeric(total)) %>% dplyr::select(state, year, total)

# Combine all
gr_total <- bind_rows(df_2019, df_2020, df_2021, df_2022, df_2023) %>% arrange(year, state)
```

## Exploratory Data Analysis

### 2.1 National Revenue Trend (2019–2023)

```{r national-trend}
gr_total %>%
  group_by(year) %>%
  summarise(national_total = sum(total, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = national_total)) +
  geom_line(size = 1.2, colour = "steelblue") +
  geom_point(size = 3) +
  labs(title = "National IGR Trend (2019–2023)", y = "Total IGR (₦)", x = "Year") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```

### 2.2 State-by-State IGR Overview

```{r allstates-bar-each-year}
gr_total %>%
  ggplot(aes(x = reorder(state, total), y = total, fill = state)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ year, scales = "free") +
  labs(title = "All States by IGR for Each Year", x = "State", y = "Total IGR (₦)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```

### 2.3 Yearly Distribution of State Revenues

```{r igr-distribution}
gr_total %>%
  ggplot(aes(x = total)) +
  geom_histogram(fill = "dodgerblue", bins = 30, colour = "white") +
  facet_wrap(~ year, scales = "free") +
  labs(title = "Distribution of IGR Across States by Year", x = "Total IGR (₦)", y = "Count of States") +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal()
```

### 2.4 Correlation Between States Over Time

```{r correlation-heatmap-clean}
# Prepare matrix (years = rows, states = columns)
state_matrix <- gr_total %>%
  pivot_wider(names_from = state, values_from = total) %>%
  column_to_rownames("year") %>%
  as.data.frame()

# Correlation matrix
cor_states <- cor(state_matrix, use = "pairwise.complete.obs")

# Abbreviate state names to 3 letters
colnames(cor_states) <- substr(colnames(cor_states), 1, 3)
rownames(cor_states) <- substr(rownames(cor_states), 1, 3)

# Plot clean heatmap with larger tiles
corrplot(cor_states,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.cex = 0.5,            # slightly larger text labels
         col.lim = c(-1, 1),
         cl.ratio = 0.50,         # slightly wider color legend
         cl.align.text = 'l',
         title = "Correlation Heatmap of IGR Between States (2019–2023)",
         mar = c(1, 1, 2, 1),     # more padding around the plot
         addgrid.col = NA)        # remove grid lines for cleaner tiles
```

## Clustering Analysis

### 3.1 Identifying Revenue-Based Clusters

```{r clustering-states}
# Step 1: Compute features (mean and growth)
gr_summary <- gr_total %>%
  group_by(state) %>%
  summarise(
    avg_igr = mean(total, na.rm = TRUE),
    growth = (total[year == 2023] - total[year == 2019]) / total[year == 2019]
  ) %>%
  drop_na()

# Step 2: Standardise data
scaled_features <- scale(gr_summary[, c("avg_igr", "growth")])

# Step 3: Elbow method to choose optimal k
fviz_nbclust(scaled_features, kmeans, method = "wss") +
  labs(
    title = "Elbow Method: Optimal Number of Clusters",
    subtitle = "The bend (elbow) at k = 3 suggests this is the ideal number of clusters",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Sum of Squares"
  ) +
  theme_minimal()

# Step 4: K-means clustering (k = 3 for now)
k_res <- kmeans(scaled_features, centers = 3, nstart = 25)
gr_summary$cluster <- as.factor(k_res$cluster)

# Step 5: Visualise clusters with clearer labels

# Add cluster labels and scaled values to gr_summary
gr_summary <- gr_summary %>%
  mutate(cluster = as.factor(k_res$cluster),
         avg_igr_scaled = scaled_features[, "avg_igr"],
         growth_scaled = scaled_features[, "growth"])

# Custom ggplot cluster visualisation
ggplot(gr_summary, aes(x = avg_igr_scaled, y = growth_scaled, colour = cluster, label = state)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(show.legend = FALSE, size = 5.5, max.overlaps = 50, force = 2) +
  labs(
    title = "Clusters of Nigerian States by Avg IGR & Growth",
    subtitle = "Clustered based on average IGR and 5-year growth trend",
    x = "Scaled Average IGR",
    y = "Scaled Growth Rate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### 3.2 Statistical Comparison Between Clusters

```{r anova-tukey-test}
# ANOVA: Is avg IGR different across clusters?
anova_model <- aov(avg_igr ~ cluster, data = gr_summary)
summary(anova_model)

# Post-hoc test: Tukey HSD
TukeyHSD(anova_model)

# Visualise differences in avg IGR
ggboxplot(gr_summary, x = "cluster", y = "avg_igr", fill = "cluster", palette = "jco") +
  labs(title = "Comparison of Average IGR Across Clusters",
       x = "Cluster", y = "Average IGR (₦)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```

## Forecasting National IGR (2024–2025)

```{r ets-national-forecast}
# Step 1: Aggregate national total IGR by year
national_igr <- gr_total %>%
  group_by(year) %>%
  summarise(total_igr = sum(total, na.rm = TRUE))

# Step 2: Convert to time series object
ts_national <- ts(national_igr$total_igr, start = 2019, frequency = 1)

# Step 3: Fit ETS model and forecast
ts_model <- ets(ts_national)
ts_forecast <- forecast(ts_model, h = 2)

# Step 4: Create combined data frame with actual + forecast
forecast_df <- data.frame(
  year = 2019:2025,
  total_igr = c(as.numeric(ts_national), as.numeric(ts_forecast$mean)),
  lower = c(rep(NA, length(ts_national)), ts_forecast$lower[,2]),
  upper = c(rep(NA, length(ts_national)), ts_forecast$upper[,2]),
  type = c(rep("actual", length(ts_national)), rep("forecast", 2))
)

# Plot with continuous line and separate point colours
ggplot(forecast_df, aes(x = year, y = total_igr)) +
  geom_line(size = 1, colour = "steelblue") +  # continuous line
  geom_point(aes(colour = type), size = 3) +
  geom_ribbon(
    data = subset(forecast_df, type == "forecast"),
    aes(ymin = lower, ymax = upper),
    fill = "cyan", alpha = 0.2
  ) +
  scale_colour_manual(values = c("actual" = "tomato", "forecast" = "dodgerblue")) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = 2019:2025) +
  labs(
    title = "ETS Forecast of National IGR (2019–2025)",
    x = "Year",
    y = "Total IGR (₦)",
    colour = "Type"
  ) +
  theme_minimal()
```
